Double White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-07-20 papers/wdmerger_I/paper.tex
Add a first cut at the section on the circular orbit test.

%\documentclass[12pt,preprint]{../aastex}
\documentclass[iop]{../emulateapj}
% for \sout
\usepackage{ulem}
% makes sure \em{} is italic rather than underlined (corrects ulem from line above)
\normalem

@@

\newcommand{\rhocutoff}{\rho_\mathrm{cutoff}}
\newcommand{\rhoanelastic}{\rho_\mathrm{anelastic}}

\newcommand{\gcc}{\mathrm{g~cm^{-3} }}
\newcommand{\Tcutoff}{T_\mathrm{cutoff}}

@@

should show:

effect of resolution

inertial vs. rotating frame

effect of ppm\_reference and gravity update type

effect of boundary conditions (size of domain and possibly multipole dependence)

@@

We presented a set of numerical tests that show where we can 
and cannot trust these techniques. We spent much time considering 
the role of bulk motions on the grid and we conclude that there 
are real issues with substantial bulk velocities on static grids, 
that can diminish the quality of the resulting solutions to the 
fluid equations. However, these effects diminish with increasing 
resolution. We therefore make no explicit claims about the usefulness 
of Langrangian versus Eulerian methods and instead simply observe 
that whenever a simulation is performed, it is important to have 
a measure of numerical accuracy and a sense of whether we are 
witnessing properties that converge with increasing resolution.
These problems do suggest that, where possible, we should seek to 
minimize bulk motions on static grids. A comparison of orbit 
simulations in both rotating and inertial reference frames 
demonstrates that in practice this is not so simple, and that 
a rotating reference frame has its own numerical issues for 
the type of simulation we desire to perform here. Perhaps contrary 
to expectations, the stars appear to be more stable in an inertial 
frame where they are constantly in motion on the grid. It 
is not easy to predict the correct behavior of such systems, 
so the best recourse for assessing confidence in a model from a 
verification standpoint is still to see whether the observed 
behavior converges with resolution.

>>>>

\documentclass[iop]{../emulateapj}

@@

@@

We now consider the phase of the binary system where the stars are orbiting each other 
at distances great enough that the initial orbits should be approximately Keplerian. 
There are a number of effects worth looking into here. For simplicity, we choose two 
cases to demonstrate the simulation behavior: an equal mass case of two $0.9\msolar$ 
white dwarfs, and an unequal mass case of $0.9\msolar$ and $0.6\msolar$ white dwarfs.
In each case, the initial orbital period is 100 seconds.

For some of the algorithms described earlier in this work, a single orbit of these 
systems is enough to examine their effects. In Section \ref{sec:gravity_boundary_conditions},
we discussed the replacement of a monopole boundary condition solver for the gravitational 
potential with a more general multipole solver for the boundaries. To test the relevance 
of this effect, we considered a single orbit of the unequal mass system and measured 
the distance between the two white dwarfs at the beginning of the simulation and after 
the full orbital period. This distance should not change significantly over that timescale.
We performed this test for maximum multipole moments ranging from 0 (the monopole term) to 16.
The results are shown in Figure \ref{fig:gravity_bcs}. Terms in the boundary potential 
that vary faster than $r^{-5}$ are effectively negligible in determining the outcome of the orbit, 
justifying our typical choice of maintaining terms up to $r^{-7}$.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/gravity_bcs}
  \caption{Absolute magnitude of the relative change in the distance of two unequal mass white dwarfs after one orbital period. 
           The stars were evolved in an inertial reference frame. The horizontal axis is the number of terms or multipole moments 
           captured in the series expansion for the potential at the domain boundary.\label{fig:gravity_bcs}}
\end{figure}

Another that we consider is the energy conservation of the system. Recalling Section \ref{sec:gravity_hydro_coupling}, 
there are several different methods of applying the gravitational source term to the hydrodynamics equations. In CASTRO we presently 
have four options, controlled by the parameter {\tt castro.grav\_source\_type}, which we will shortern to {\tt gs} for the present 
discussion. ${\tt gs} = 1$ is the standard cell-centered source term for gravity. ${\tt gs} = 2$ is a slight variation on this scheme 
that determines the value of the energy source term after the momentum source term has been applied; the former approach uses 
the uncorrected momenta in calculating $\rho \mathbf{u} g$, which we have found to be more accurate. ${\tt gs} = 3$ is entirely 
different: after calculating the new momenta, we reset the total energy to be equal to the internal energy plus the 
kinetic energy. This has the virtue of ensuring that there is no conflict due to discretization between the momentum and 
energy equations, and also correctly ensuring that the gravitational force does not directly change the internal energy -- 
and thus the temperature -- of the fluid. However, it explicitly sacrifices total energy conservation. ${\tt gs} = 4$ is 
the new conservative method of evaluating the energy source terms at cell faces. The results for the change in energy 
after a single orbit are seen in the first column of Table \ref{table:sources}. The first two versions give reasonable and 
similar levels of energy conservation. The third has total energy changes on the order of 100\%, but this itself 
does not have a severe effect on the dynamics because in this scheme the total energy variable is effectively a placeholder 
value of the kinetic energy plus internal energy, rather than being evolved directly. The last scheme is nearly two orders 
of magnitude better in energy conservation, justifying the effort in varying the scheme.

In Table \ref{table:sources} we show also the effects on energy conservation of using the inertial reference frame. 
We use ${\tt rs}$ for the CASTRO parameter {\tt castro.rot\_source\_type}, and we have the same four options for updating
the energy source term as for gravity. The ${\tt rs} = 0$ column means that rotation is turned off and we are 
in the inertial frame. We see that the choice of rotational coupling is much less important than the choice of gravity coupling. 
Interestingly, the ``conservative'' ${\tt rs} = 4$ is slightly worse in energy conservation than the non-conservative, 
cell-centered ${\tt rs} = 2$ algorithm. This may be because the rotational force grows with distance, and so losses at the 
domain boundaries may be enhanced by explicitly evaluating the flux at cell edges.

\input{plots/sources.table}

We are most interested in the stability of these systems over long timescales. To this end, we consider the same systems as above, 
but evolve them for 25 orbital periods. In Figure \ref{fig:circular_orbit_comparison} we illustrate the evolution of these systems 
by plotting the center of mass locations of the white dwarfs on the orbital ($xy$) plane. For the equal mass case in the inertial 
reference frame, the curves essentially fall on top of each other, indicating that the stars are indeed orbiting at the initial 
distance without substantial deviation. In the unequal mass case, the stars now orbit at different distances but do not much 
deviate from their initial distance. We do observe a secular decline in the center of mass of these stars, indicating that 
eventually the stellar orbits will degrade due to numerical error even without intervention. The co-rotating frame is different.
For clarity of visualization, we rotate these results back into the inertial frame before displaying their orbits.
The stellar orbit in the equal mass case does not show the same secular decline, and the system appears to be stable. However, 
the cost is that the oscillations about this equilibrium are slightly larger than in the inertial frame. The unequal mass case displays 
this effect starkly: while the average distance is not declining, there are huge oscillations about this average. In addition, 
the stars move substantially within the rotating frame itself, yielding the chaotic-looking orbits in the figure. The oscillations 
grow with time and will eventually cause the stars to merge; however, the stars would eventually merge even in the inertial frame 
case, and at a faster rate than the equal mass case. This suggests that this mass ratio is too large to ignore tidal effects, and 
that the tidal effects play out much differently in the rotating frame.

\begin{figure*}
  \centering
  \includegraphics[scale=0.7]{plots/circular_orbit_comparison}
  \caption{Positions of the white dwarfs in the orbital plane for four cases evolved over 25 orbital periods. 
           The $x$ and $y$ axes are normalized to the size of the domain, so that $x = -0.5$ is the left edge 
           and $x = 0.5$ is the right edge. The dashed blue curve is the position of the primary white dwarf, 
           and the solid red curve is the position of the secondary. In plot (a) we have the equal mass system 
           evolved in the inertial reference frame, and in plot (c) we have the same system evolved in a rotating frame, 
           where the positions have been transformed back to the inertial frame for comparison. 
           Plots (b) and (d) are analogous but for the unequal mass system.\label{fig:circular_orbit_comparison}}
\end{figure*}

Turning to the conservation properties of the system, we examine as fairly typical cases the equal mass system 
in the inertial frame for energy conservation (Figure \ref{fig:energy_conservation_equal}), and the unequal mass 
system in the rotating frame for angular momentum conservation (Figure \ref {fig:angular_momentum_conservation_unequal}). 
Angular momentum is conserved to within a few percent over the 25 orbits, while energy conservation is about 
an order of magnitude better. We note that while this is already a fairly good level of energy conservation, it is not 
nearly as good as the results of \citet{marcello:2012}. This is because we reset the internal energy to a level 
corresponding to our temperature floor when it goes negative, while \citeauthor{marcello:2012} do not reset and 
instead ignore the internal energy if it is negative. The resets impose an artificial floor on our ability to 
conserve energy, but they only happen in low-density regions and do not much affect the large-scale dynamics. 
Meanwhile, angular momentum conservation is not quite as good. This is linked to the decline (or increase) in 
the size of the orbit. This implies that we ought to be careful in concluding that at these moderate resolutions 
we can safely evolve systems for many dozens of orbits; this needs to be verified to ensure that an observed 
inspiral and merger is physically (not numerically) motivated.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/equal_energy_rot_0.eps}
  \caption{Absolute magnitude of the relative change in energy of two equal mass white dwarfs through 25 orbital periods,
           evolved in an inertial reference frame. The decline and recovery is a change in sign of the energy difference.
           \label{fig:energy_conservation_equal}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/unequal_angular_momentum_rot_1.eps}
  \caption{Absolute magnitude of the relative change in angular momentum of two equal mass white dwarfs after 25 orbital periods, 
           evolved in a co-rotating reference frame. We consider only the component of the angular moment along the rotational 
           axis. The decline and recovery is a change in sign of the angular momentum difference.
           \label{fig:angular_momentum_conservation_unequal}}
\end{figure}

@@

We presented a set of numerical tests that show where we can 
and cannot trust these techniques. We spent much time considering 
the role of bulk motions on the grid and we conclude that there 
are real issues with substantial bulk velocities on static grids, 
that can diminish the quality of the resulting solutions to the 
fluid equations. However, these effects diminish with increasing 
resolution. We therefore make no explicit claims about the usefulness 
of Langrangian versus Eulerian methods and instead simply observe 
that whenever a simulation is performed, it is important to have 
a measure of numerical accuracy and a sense of whether we are 
witnessing properties that converge with increasing resolution.
These problems do suggest that, where possible, we should seek to 
minimize bulk motions on static grids. A comparison of orbit 
simulations in both rotating and inertial reference frames 
demonstrates that in practice this is not so simple, and that 
a rotating reference frame has its own numerical issues for 
the type of simulation we desire to perform here. Based on 
the tests performed here, in some cases the stars may be 
\textit{more} stable in an inertial frame despite 
constantly being in motion on the grid. It 
is not easy to predict the correct behavior of such systems, 
though one recourse for assessing confidence in a model from a 
verification standpoint is to see whether the observed 
behavior converges with resolution. However, it is not clear 
whether at practical resolutions the results in the 
rotating and inertial frames will converge to each other.
